\documentclass[12pt,preprint]{article}

\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[top=1in, bottom=1.in, left=1in, right=1in]{geometry}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[version=4]{mhchem}
\usepackage[numbers,super,sort&compress]{natbib}
\usepackage{natmove}
\usepackage[newfloat]{minted}
\usepackage[linktocpage,pdfstartview=FitH,colorlinks,
linkcolor=blue,anchorcolor=blue,
citecolor=blue,filecolor=blue,menucolor=blue,urlcolor=blue]{hyperref}
\usepackage{amsmath,amssymb,algorithm,algpseudocode}
\usepackage{booktabs,multirow,graphicx}

\title{discopt: A Differentiable Mixed-Integer Nonlinear Programming Solver with a Memory-Safe Interior Point Method}
\author{John R. Kitchin}
\date{2026-02-14}

\begin{document}

\maketitle

\section{Abstract}
\label{org-abstract}
We present discopt, a hybrid Rust/JAX/Python solver for mixed-integer nonlinear programming (MINLP) problems, together with ripopt, a faithful Rust translation of Ipopt's interior point algorithm. The system makes five contributions: (1) differentiable MINLP solving via implicit differentiation of the KKT system, enabling decision-focused learning through nonconvex optimization layers; (2) batched node evaluation using \texttt{jax.vmap} for data-parallel branch-and-bound; (3) a pure-JAX interior point method with Mehrotra predictor-corrector steps and Ipopt-style filter line search that is JIT-compilable and vmappable; (4) ripopt, a memory-safe Rust IPM with zero C/Fortran dependencies that originated from Ipopt's codebase and extends it with hybrid fallback strategies, automatic problem reformulation, and native Rust linear algebra; and (5) a compositional relaxation compiler supporting 25 McCormick relaxation functions, $\alpha$ BB, piecewise McCormick, specialized envelopes, and ICNN-learned relaxations. The modeling API supports generalized disjunctive programming (GDP) with automatic big-M and convex hull reformulations, indicator constraints, and logical propositions. The solver includes an expression DAG convexity detector that enables adaptive dispatch (skipping relaxation overhead for convex problems), per-constraint outer approximation cuts, reliability branching with pseudocost tracking and LP-based strong branching, and OBBT with incumbent cutoff for progressive bound tightening. All three NLP backends (pure-JAX IPM, ripopt, Ipopt) produce identical optimal solutions on a suite of 9 NLP backend comparison problems and 24 correctness problems (12 NLP + 12 MINLP) with zero incorrect results. On a decision-focused learning task, differentiating through the solver achieves 36\% regret reduction versus two-stage predict-then-optimize. The algebraic coefficient extraction path provides 209--2539$\times$ speedups over autodiff-based extraction for LP and QP subproblems. Source code is available under EPL-2.0 license (discopt and ripopt).


\section{Introduction}
\label{org-introduction}
Mixed-integer nonlinear programming (MINLP) is a powerful framework for optimization problems involving both discrete choices and nonlinear relationships, with applications spanning process synthesis, network design, scheduling, and machine learning pipeline optimization \cite{Belotti2013,Grossmann2021}. The dominant algorithmic paradigm for deterministic global MINLP is spatial branch and bound (sBB), which enumerates a search tree of progressively tighter variable-bound regions, solving a convex relaxation at each node \cite{Land1960,Dakin1965,Smith1999}.


State-of-the-art MINLP solvers---BARON \cite{Tawarmalani2005}, Couenne \cite{Belotti2009couenne}, SCIP \cite{Bestuzheva2023}, and Bonmin \cite{Bonami2008} are monolithic C, C++, or Fortran codebases that have accumulated decades of numerical hardening. While effective, this architecture creates three gaps in the current solver ecosystem.


This paper describes discopt and ripopt, which together address these three gaps. The contributions are:


\begin{enumerate}
\item \textbf{Differentiable MINLP} via two mechanisms: the envelope theorem (Level 1) for fast objective sensitivity, and implicit differentiation of the KKT system (Level 3) for full primal--dual sensitivity matrices, enabling decision-focused learning through nonconvex solvers (\S\ref{org-differentiable-optimization}).

\item \textbf{Batched node evaluation} using \texttt{jax.vmap} over NLP relaxation solves at B\&B nodes, with warm-starting from parent solutions (\S\ref{org-batch-solving-via-jax-vmap}).

\item \textbf{Pure-JAX interior point method} with Mehrotra predictor-corrector \cite{Mehrotra1992}, Cholesky-based inertia correction, Ipopt-style filter line search \cite{Wachter2006}, and carry-based \texttt{jax.lax.while\_loop} for JIT cache reuse (\S\ref{sec-jax-ipm}).

\item \textbf{ripopt}: a memory-safe Rust NLP solver that originated as a source-level translation of Ipopt's core IPM and has been extended with hybrid fallback strategies, automatic problem reformulation, and native Rust linear algebra (\S\ref{sec-ripopt}).

\item \textbf{Compositional relaxation framework} with 25 McCormick relaxation functions \cite{McCormick1976,Tsoukalas2014}, $\alpha$ BB underestimators \cite{Androulakis1995,Adjiman1998}, piecewise McCormick with adaptive partitioning \cite{Castro2015} achieving 97--98\% gap reduction, and ICNN-learned relaxations \cite{Amos2017icnn} (\S\ref{org-convex-relaxation-framework}).

\end{enumerate}


The remainder of this paper is organized as follows. Section~\ref{org-background-and-related-work} reviews background and related work. Section~\ref{org-the-discopt-architecture} describes the discopt architecture. Section~\ref{org-nlp-solver-backends} presents the three NLP solver backends. Section~\ref{org-convex-relaxation-framework} details the convex relaxation framework. Section~\ref{org-differentiable-optimization} covers differentiable optimization. Section~\ref{org-computational-experiments} reports computational experiments. Section~\ref{org-discussion} discusses limitations and future work, and Section~\ref{org-conclusion} concludes.


\section{Background and Related Work}
\label{org-background-and-related-work}
\subsection{Spatial Branch and Bound for MINLP}
\label{org-spatial-branch-and-bound-for-minlp}
Spatial branch and bound (sBB) is the dominant paradigm for deterministic global optimization of MINLP problems \cite{Belotti2013}. The algorithm maintains a search tree where each node represents a subproblem defined by tightened variable bounds. At each node, a convex relaxation of the original nonconvex problem is solved; the relaxation's optimal value provides a valid lower bound on the global optimum within that region.


The framework was introduced by Land and Doig \cite{Land1960} for integer programming and extended by Dakin \cite{Dakin1965}. Smith and Pantelides \cite{Smith1999} formalized spatial B\&B for nonconvex MINLP by combining variable-bound branching with factorable relaxation construction.


Modern sBB solvers enhance the basic framework with bound tightening techniques---feasibility-based bound tightening (FBBT) and optimization-based bound tightening (OBBT) \cite{Belotti2009,Gleixner2017} --- which propagate bound improvements through the expression graph, and primal heuristics such as the feasibility pump \cite{Fischetti2005} that attempt to find good incumbent solutions early in the search.

\subsection{Convex Relaxations}
\label{org-convex-relaxations}
The quality of the convex relaxation at each node directly determines the effectiveness of the B\&B search. Weaker relaxations produce loose lower bounds, leading to more nodes explored before convergence.


McCormick relaxations \cite{McCormick1976} construct convex and concave envelopes for factorable functions by composing relaxations of elementary operations (addition, multiplication, univariate functions). Tsoukalas and Mitsos \cite{Tsoukalas2014} extended this to the multivariate case. The $\alpha$ BB method \cite{Androulakis1995,Adjiman1998} adds a separable quadratic perturbation to ensure convexity, where the perturbation magnitude $\alpha$ is determined by the minimum eigenvalue (or Gershgorin estimate) of the objective Hessian.


Piecewise McCormick relaxations \cite{Castro2015} partition the variable domain into subintervals and construct tighter McCormick envelopes on each partition, significantly reducing the relaxation gap at the cost of additional auxiliary variables. Outer approximation (OA) \cite{Duran1986,Fletcher1994} and reformulation-linearization technique (RLT) \cite{Sherali1990} provide complementary cutting-plane approaches.

\subsection{Interior Point Methods for NLP}
\label{org-interior-point-methods-for-nlp}
The continuous NLP relaxation at each B\&B node is typically solved by an interior point method (IPM). The primal-dual IPM \cite{Wright1997,Nocedal2006} transforms inequality constraints into barrier terms in the objective, then applies Newton's method to the resulting KKT system. Key algorithmic components include the Mehrotra predictor-corrector step \cite{Mehrotra1992}, which adaptively adjusts the centering parameter for faster convergence, and filter line search \cite{Wachter2005,Wachter2006}, which accepts steps that improve either the objective or constraint feasibility.


Ipopt \cite{Wachter2006} is the most widely used open-source NLP solver, implementing a filter line-search interior point method with inertia correction, feasibility restoration, and warm-starting. It has accumulated over 20 years of numerical hardening for edge cases including degenerate Jacobians, rank-deficient KKT systems, and cycling detection.


Recent work on JAX-native optimization includes MPAX \cite{Lu2024}, a first-order solver for large-scale LP and QP that implements restarted primal-dual hybrid gradient (PDHG) variants: r2HPDHG for LP and rAPDHG for QP, with Ruiz and Pock-Chambolle diagonal preconditioning, adaptive restarts, and infeasibility detection. MPAX leverages JAX for hardware portability (CPU/GPU/TPU), batched solving via \texttt{vmap}, multi-GPU distribution via \texttt{pmap}, and automatic differentiation through solutions. The two systems occupy complementary niches: MPAX's first-order iterations are cheap (one matrix-vector product each) but converge slowly ($O(1/k)$ to $O(1/k^2)$ with restarts), making it well-suited for very large sparse LP/QP instances ($n > 10^5$) where moderate accuracy suffices. discopt's second-order IPM performs expensive Cholesky factorizations ($O(n^3)$) but converges quadratically, targeting the small-to-medium nonconvex NLP subproblems ($n < 10^4$) that arise at each B\&B node where high-accuracy solutions are needed for valid dual bounds. Unlike MPAX, discopt handles nonconvex objectives (via inertia correction), nonlinear constraints (via augmented KKT), and integer variables (via spatial B\&B). JAX device placement in discopt is controlled by the \texttt{JAX\_PLATFORMS} environment variable, inheriting GPU/TPU acceleration automatically when a suitable backend is available.

Sensitivity analysis of NLP solutions with respect to problem parameters is well-established \cite{Fiacco1983}. Pirnay et al. \cite{Pirnay2012} implemented the sIPOPT framework within Ipopt for efficient parametric sensitivity computation by reusing the KKT factorization.

\subsection{Differentiable Optimization}
\label{org-differentiable-optimization-background}
The idea of differentiating through an optimization solver to enable end-to-end learning has received significant recent attention. Amos and Kolter \cite{Amos2017} introduced OptNet for differentiating through quadratic programs. Agrawal et al. \cite{Agrawal2019} extended this to general convex programs via cvxpylayers. Elmachtoub and Grigas \cite{Elmachtoub2022} formalized the "smart predict, then optimize" framework, and Wilder et al. \cite{Wilder2019} and Mandi et al. \cite{Mandi2020} developed methods for combinatorial optimization with decision-focused learning.


DiffOpt.jl \cite{DiffOpt2025} extends differentiable optimization in the Julia/JuMP ecosystem to smooth, potentially nonconvex NLPs via KKT differentiation at locally optimal solutions, but explicitly excludes mixed-integer formulations: ``extending robustness to nonsmooth regimes, mixed-integer formulations, and broader conic-nonlinear composites remains an open direction'' \cite{DiffOpt2025}.

A critical gap remains: \textbf{none of these approaches handle nonconvex MINLP}. Existing differentiable optimization layers require convexity (OptNet, cvxpylayers), handle only continuous NLP (DiffOpt.jl), or use surrogate loss functions that approximate but do not exactly differentiate through the solver (decision-focused learning for combinatorics). discopt fills this gap by providing exact gradients through a nonconvex NLP solver via implicit differentiation of the KKT conditions.

\subsection{Gaps in the State of the Art}
\label{org-gaps-in-the-state-of-the-art}
Table \ref{tab-gaps} summarizes the capabilities of existing systems versus discopt.


\begin{table}[htbp]
\centering
\begin{tabular}{llllllll}
\toprule
Feature & BARON & SCIP & Bonmin & cvxpylayers & DiffOpt.jl & MPAX & discopt \\
\midrule
Nonconvex MINLP & Yes & Yes & Yes & No & No & No & Yes \\
Nonlinear constraints & Yes & Yes & Yes & No & Yes & No & Yes \\
GDP modeling & No & Partial & No & No & No & No & Yes \\
Differentiable solve & No & No & No & Yes (convex) & Yes (NLP) & Yes (LP/QP) & Yes \\
Batched solving & No & No & No & N/A & No & Yes & Yes \\
GPU/TPU support & No & No & No & N/A & No & Yes & Yes \\
Memory-safe solver & No & No & No & N/A & N/A & Yes & Yes \\
Zero C/Fortran deps & No & No & No & Yes & N/A & Yes & Yes \\
DFL through MINLP & No & No & No & No & No & No & Yes \\
Compositional relaxation & Yes & Partial & No & N/A & N/A & N/A & Yes \\
\bottomrule
\end{tabular}
\caption{Comparison of discopt with existing optimization frameworks. DFL = decision-focused learning. GPU/TPU support for discopt is via JAX backend selection (\texttt{JAX\_PLATFORMS}). DiffOpt.jl differentiates through smooth NLP but excludes mixed-integer formulations.}
\label{tab-gaps}
\end{table}


\section{The discopt Architecture}
\label{org-the-discopt-architecture}
\subsection{System Overview}
\label{org-system-overview}
discopt is organized as a three-layer system: (1) a Python modeling API for problem formulation, (2) a JAX-based numerical layer for evaluation, automatic differentiation, and batch computation, and (3) a Rust backend for tree management, expression structure detection, and presolve.


\begin{minted}[frame=lines,fontsize=\scriptsize,linenos]{text}
  User Model (Python)
       |
       v
  Expression DAG ---> DAG Compiler ---> JAX Functions (JIT/grad/vmap)
       |                                        |
       v                                        v
  Rust Expression IR             NLP Evaluator (obj, grad, Hess, Jac)
       |                                        |
       v                                        v
  B&B Tree Manager               NLP Solver Backend
  (node pool, branching,         (IPM / ripopt / Ipopt)
   bound tightening)                    |
       |                                v
       +-------- batch dispatch --------+
                     |
                     v
              SolveResult
\end{minted}

An automatic problem classifier inspects the expression DAG to categorize problems as LP, QP, NLP, MILP, MIQP, or MINLP. For LP and QP subclasses, discopt dispatches to specialized IPM solvers that extract algebraic coefficients directly from the DAG, bypassing JAX tracing and autodiff entirely. This algebraic extraction path provides 209--2539$\times$ speedups over autodiff-based extraction (\S\ref{org-performance}).

\subsection{Modeling API}
\label{org-modeling-api}
The modeling API uses operator overloading to construct an expression DAG that maps to both JAX-traceable functions and Rust expression IR:


\begin{minted}[frame=lines,fontsize=\scriptsize,linenos]{python}
import discopt.modeling as dm

m = dm.Model("portfolio")
x = m.continuous("weights", shape=(n,), lb=0, ub=1)
y = m.binary("selected", shape=(n,))

m.minimize(risk @ (x * x) - expected_return @ x)
m.subject_to(dm.sum(x) == 1, name="budget")
m.subject_to(x <= y, name="indicator")  # x_i > 0 => y_i = 1
m.subject_to(dm.sum(y) <= k, name="cardinality")

result = m.solve()
\end{minted}

The API supports continuous, binary, and integer variables; parameters (for sensitivity analysis and decision-focused learning); generalized disjunctive programming (GDP) constraints with automatic big-M or convex hull reformulation (\S\ref{sec-gdp}); logical propositions; and standard mathematical functions (\texttt{exp}, \texttt{log}, \texttt{sin}, \texttt{cos}, \texttt{sqrt}, \texttt{abs}, and matrix operations).

\subsection{JAX DAG Compiler}
\label{org-jax-dag-compiler}
The DAG compiler walks the expression tree and produces pure JAX functions compatible with \texttt{jax.jit}, \texttt{jax.grad}, and \texttt{jax.vmap}. Each expression node type (\texttt{Variable}, \texttt{Constant}, \texttt{BinaryOp}, \texttt{FunctionCall}, etc.) maps to a corresponding JAX operation. The compiler handles:


\begin{itemize}
\item \textbf{Lazy evaluation}: expressions are not evaluated until the DAG is compiled

\item \textbf{Automatic differentiation}: the JAX function graph supports forward and reverse mode AD for gradients, Hessians, and Jacobians

\item \textbf{Vectorization}: \texttt{jax.vmap} can map the compiled function over batches of variable values or bound vectors

\item \textbf{JIT compilation}: compiled functions are traced and optimized by XLA for efficient repeated evaluation

\end{itemize}

\subsection{Rust Backend}
\label{org-rust-backend}
The Rust backend (\texttt{discopt-core} and \texttt{discopt-python} crates) provides three components: (1) an expression IR that mirrors the Python DAG, (2) a B\&B tree manager with node pool, branching, and pseudocost tracking, and (3) presolve routines including FBBT, probing, and simplification. All Rust data structures are exposed to Python via PyO3 bindings \cite{pyo3} with zero-copy numpy array transfer.

\subsubsection{Reliability Branching}
\label{org-reliability-branching}
Variable selection in B\&B significantly impacts solver performance. discopt implements reliability branching \cite{Achterberg2005}, which combines pseudocost branching with strong branching in a hybrid scheme.

After each branching decision, the Rust \texttt{TreeManager} records the bound improvement per unit of variable change (pseudocost) for both up and down branches. Running averages are maintained per variable. When selecting a branching variable, the algorithm examines all fractional integer variables and:

\begin{enumerate}
\item For variables with $\ge \eta_{\text{rel}}$ (default 8) pseudocost observations: scores are computed as the product $(10^{-6} + d_i^-)(10^{-6} + d_i^+)$ where $d_i^{\pm}$ are the pseudocost estimates for down/up branching.
\item For variables with $< \eta_{\text{rel}}$ observations: Python-side strong branching via LP relaxation solves with HiGHS \cite{Huangfu2018}. For each candidate, two LP subproblems are constructed by linearizing the NLP at the current solution (gradient as objective, Jacobian-based constraint linearization), one with the variable's upper bound fixed to $\lfloor x_i \rfloor$ (down) and one with the lower bound fixed to $\lceil x_i \rceil$ (up). The resulting LP bound improvements determine the score.
\item Branch hints are passed from Python back to the Rust \texttt{TreeManager} via \texttt{set\_branch\_hints()}, where they override the default most-fractional selection.
\end{enumerate}

As the tree matures and pseudocost observations accumulate, the algorithm naturally transitions from expensive strong branching to cheap pseudocost branching, achieving near-optimal variable selection at amortized low cost.

\subsubsection{OBBT with Incumbent Cutoff}
\label{org-obbt-incumbent-cutoff}
Optimality-based bound tightening (OBBT) \cite{Gleixner2017} solves LP relaxations to find the tightest possible bounds on each variable: $\underline{x}_i = \min_{x} x_i$ and $\overline{x}_i = \max_{x} x_i$ subject to the model's linear constraints and variable bounds.

When an incumbent solution with objective value $z^*$ is known, discopt adds the constraint $c^T x \le z^*$ (using linear objective coefficients) to the OBBT LPs. This often dramatically tightens bounds because many variable values become infeasible when the objective must improve upon the incumbent. For maximization problems, the constraint is negated: $-c^T x \le -z^*$.

OBBT is invoked periodically during B\&B whenever a new incumbent is found and the problem has $\le 200$ variables. Each LP is solved with HiGHS using warm-starting from the previous LP's basis, keeping the per-variable LP solve cost low. Tightened bounds propagate to all open nodes, potentially enabling additional pruning and tighter relaxations at subsequent nodes.

\subsubsection{FBBT with Incumbent Cutoff}
\label{org-fbbt-incumbent-cutoff}
Feasibility-based bound tightening (FBBT) \cite{Belotti2009,Ryoo1996} propagates bound information through the expression DAG without solving auxiliary optimization problems, making it much cheaper than OBBT. Each constraint $g(x) \le 0$ induces range implications: if a variable's contribution to $g$ exceeds the slack, its domain can be tightened.

When an incumbent with objective value $z^*$ is available, discopt adds the cutoff constraint $f(x) \le z^*$ to the FBBT propagation. This is implemented in the Rust \texttt{fbbt\_with\_cutoff} function, which takes the expression IR, current variable bounds, and the incumbent objective. The cutoff often triggers cascading domain reductions: if reducing one variable's range via the objective cutoff renders a constraint infeasible for extreme values of another variable, that variable's bounds tighten as well.

FBBT with incumbent cutoff runs when a new incumbent is found, while OBBT is reserved for periodic invocation on problems with $\le 200$ variables. The combination provides complementary bound tightening: FBBT is fast but conservative; OBBT is expensive but finds the tightest LP-based bounds.

\subsection{Generalized Disjunctive Programming}
\label{sec-gdp}

Generalized disjunctive programming (GDP) \cite{Raman1994,Grossmann2002} provides a higher-level modeling abstraction for optimization problems involving logical decisions. Rather than manually introducing big-M constraints, the modeler expresses logical structure directly, and the solver reformulates it into standard MINLP constraints automatically.

GDP extends classical MINLP by representing discrete decisions as disjunctions:
\begin{equation}
\bigvee_{k \in D_i} \left[ \begin{array}{c} Y_{ik} \\ g_{ik}(x) \le 0 \end{array} \right], \quad \Omega(Y) = \textsc{True}
\end{equation}
where $Y_{ik}$ are Boolean variables indicating which disjunct is active, $g_{ik}(x) \le 0$ are the constraints in disjunct $k$ of disjunction $i$, and $\Omega(Y)$ encodes logical propositions over the Boolean variables. This formulation separates the logical structure from the continuous optimization, enabling tighter reformulations than ad-hoc big-M modeling.

discopt supports three categories of GDP constraints:

\paragraph{Indicator constraints.}
The \texttt{if\_then(y, constraints)} method specifies that when binary variable $y = 1$, all listed constraints must hold; when $y = 0$, they are relaxed. The reformulation computes tight big-M values via interval arithmetic over the expression DAG, avoiding the common pitfall of overly conservative M values that weaken the LP relaxation \cite{Trespalacios2015}.

\paragraph{Disjunctive constraints.}
The \texttt{either\_or(disjuncts)} method specifies that exactly one group of constraints must hold. Two reformulation methods are available:

\begin{itemize}
\item \textbf{Big-M reformulation} \cite{Nemhauser1999}: introduces one binary selector $y_k$ per disjunct with $\sum_k y_k = 1$, and relaxes each constraint $g_j(x) \le 0$ in disjunct $k$ to $g_j(x) \le M_j(1 - y_k)$. Big-M values are computed automatically from variable bounds via interval arithmetic. This adds few variables but the LP relaxation quality depends on the tightness of M.

\item \textbf{Convex hull reformulation} \cite{Lee2001,Grossmann2002}: introduces disaggregated variables $v_{jk}$ for each original variable $x_j$ in each disjunct $k$, with aggregation constraints $x_j = \sum_k v_{jk}$, bound linking $l_k y_k \le v_{jk} \le u_k y_k$, and perspective reformulation of nonlinear constraints. For linear constraints, the hull reformulation substitutes disaggregated variables and scales constants by $y_k$. For nonlinear constraints, the perspective function $y_k \cdot g(v_k / y_k)$ is used with a small $\epsilon$-clamp on $y_k$ for numerical stability. The hull reformulation provides a tighter LP relaxation than big-M at the cost of additional variables and constraints.
\end{itemize}

\paragraph{SOS constraints and logical propositions.}
SOS Type 1 (at most one variable nonzero) and SOS Type 2 (at most two adjacent variables nonzero) constraints \cite{Beale1970} are reformulated via binary indicator variables with linking constraints. Additionally, the modeling API provides logical proposition methods: \texttt{implies(y1, y2)} for $y_1 \Rightarrow y_2$, \texttt{iff(y1, y2)} for equivalence, and cardinality constraints \texttt{at\_least}, \texttt{at\_most}, and \texttt{exactly} for counting constraints over binary variables.

The GDP reformulation is applied as a preprocessing step before the model enters the DAG compiler and solver. If no GDP constraints are present, the original model is returned unchanged with zero overhead. The following example illustrates a process synthesis problem with disjunctive constraints:

\begin{minted}[frame=lines,fontsize=\scriptsize,linenos]{python}
import discopt.modeling as dm

m = dm.Model("process_selection")
x = m.continuous("flow", shape=(3,), lb=0, ub=100)
y = m.binary("select", shape=(2,))

m.minimize(2 * x[0] + 3 * x[1] + x[2])

# Either process A or process B, not both
m.either_or([
    [x[0] + x[1] >= 50, x[2] <= 10],   # Process A
    [x[0] <= 20, x[1] + x[2] >= 40],    # Process B
], name="process_choice")

# If selected, must produce at least 30 units
m.if_then(y[0], [x[0] >= 30], name="min_production")

result = m.solve()                          # big-M (default)
result = m.solve(gdp_method="hull")         # convex hull
\end{minted}


\section{NLP Solver Backends}
\label{org-nlp-solver-backends}
discopt supports three interchangeable NLP solver backends, selectable via \texttt{Model.solve(nlp\_solver}...)=. All three implement the same interface: given an objective function $f(x)$, constraints $g(x)$, variable bounds $x_l \le x \le x_u$, and constraint bounds $g_l \le g(x) \le g_u$, find $x^*$ that minimizes $f$ subject to all constraints.

\subsection{Pure-JAX Interior Point Method}
\label{sec-jax-ipm}

The pure-JAX IPM implements a primal-dual barrier method that is fully JIT-compilable and vmappable. The algorithm solves the augmented KKT system:


\begin{equation}
\begin{bmatrix}
H + \Sigma + \delta_w I & J^T \\
J & -D - \delta_c I
\end{bmatrix}
\begin{bmatrix}
\Delta x \\
\Delta y
\end{bmatrix}
=
\begin{bmatrix}
r_x \\
r_y
\end{bmatrix}
\end{equation}


where $H = \nabla^2_{xx} \mathcal{L}$ is the Hessian of the Lagrangian, $\Sigma = \mathrm{diag}(z_l / s_l + z_u / s_u)$ is the barrier Hessian diagonal, $J$ is the constraint Jacobian, $D$ is the condensed slack diagonal for inequality constraints, and $\delta_w, \delta_c$ are inertia correction regularizations.

\subsubsection{Mehrotra Predictor-Corrector}
\label{org-mehrotra-predictor-corrector}
Following Mehrotra \cite{Mehrotra1992}, each iteration consists of two linear system solves:


\begin{enumerate}
\item \textbf{Affine predictor} ($\mu = 0$): solve the KKT system with zero centering to obtain $(\Delta x^{\mathrm{aff}}, \Delta y^{\mathrm{aff}})$. Compute the affine complementarity $\mu^{\mathrm{aff}}$.

\item \textbf{Centering parameter}: $\sigma = (\mu^{\mathrm{aff}} / \mu^{\mathrm{curr}})^3$, clipped to $[0, 1]$.

\item \textbf{Corrector}: re-solve with centering $\sigma\mu$ and cross-product correction terms $\Delta s_i \Delta z_i / s_i$.

\end{enumerate}


The corrector step provides second-order information about the central path, typically reducing the iteration count by 30--50\% compared to a single Newton step \cite{Mehrotra1992}.

\subsubsection{Inertia Correction}
\label{org-inertia-correction}
The KKT system must have correct inertia ($n$ positive, $m$ negative eigenvalues) for the Newton direction to be a descent direction. We detect incorrect inertia via Cholesky factorization: if the upper-left block $W = H + \Sigma + \delta_w I$ produces NaN in its Cholesky factor, the inertia is incorrect. The regularization $\delta_w$ is increased by a factor of 8 (up to $\delta_w^{\max} = 10^{40}$) until correct inertia is achieved. This Cholesky-based detection avoids the $O(n^3)$ eigenvalue decomposition, requiring only a single Cholesky attempt with NaN detection.

\subsubsection{Filter Line Search}
\label{org-filter-line-search}
Step acceptance follows Ipopt's filter line search \cite{Wachter2006}. The algorithm maintains an $\ell_1$ merit function $\phi(x) = f(x) + \nu \|c(x)\|_1$ with adaptive penalty parameter $\nu$ and performs backtracking with Armijo-type sufficient decrease. A stall detection mechanism triggers projected steepest-descent recovery after 3 consecutive line search failures, handling the Maratos effect near active bound constraints.

\subsubsection{Carry-Based \texttt{while\_loop}}
\label{org-carry-based-while-loop}
A critical implementation detail for JAX performance: all problem data (bounds, constraint bounds, masks) is placed in the \texttt{while\_loop} carry state as a \texttt{NamedTuple}, rather than captured in the closure. Closure-captured data forces JIT retracing for each new problem instance, while carry-based data allows JIT cache reuse. Problem dimensions are derived from array shapes (\texttt{c.shape[0]}, \texttt{b.shape[0]}) rather than stored as integers, since integers in the carry become JAX tracers.

\subsubsection{Batch Solving via \texttt{jax.vmap}}
\label{org-batch-solving-via-jax-vmap}
The \texttt{solve\_nlp\_batch} function applies \texttt{jax.vmap} over the initial point and variable bound dimensions:


\begin{minted}[frame=lines,fontsize=\scriptsize,linenos]{python}
def solve_nlp_batch(obj_fn, con_fn, x0_batch, xl_batch, xu_batch, g_l, g_u, opts):
    def _solve_single(x0, xl, xu):
        return ipm_solve(obj_fn, con_fn, x0, xl, xu, g_l, g_u, opts)
    return jax.vmap(_solve_single)(x0_batch, xl_batch, xu_batch)
\end{minted}

This solves all $k$ node relaxations simultaneously, providing approximately $k$-fold parallelism on hardware with sufficient FLOP throughput. The objective and constraint functions are shared across the batch; only the starting points and variable bounds vary per instance.


At the root node of the B\&B tree, multi-start solving uses 5 diverse starting points (midpoint, lower-quarter, upper-quarter, and 2 random) solved simultaneously via \texttt{vmap}. On non-root nodes, the solver warm-starts from the parent node's solution clipped to the child's tightened bounds.

\subsection{ripopt: A Memory-Safe IPM in Rust}
\label{sec-ripopt}

ripopt (Rust Interior Point OPTimizer) originated as a source-level translation of Ipopt's core interior point algorithm from C++ to Rust, preserving two decades of numerical hardening for edge cases (degenerate Jacobians, rank-deficient KKT systems, cycling detection, tiny-step recovery). Since the initial translation, ripopt has been substantially extended with hybrid fallback strategies, automatic problem reformulation, and native Rust linear algebra, making it a distinct solver rather than a simple port.

\subsubsection{Motivation}
\label{org-motivation}
Three considerations drove the initial design:

\begin{enumerate}
\item \textbf{Numerical hardening.} Starting from Ipopt's codebase preserves institutional knowledge accumulated over two decades, avoiding the need to rediscover edge cases one benchmark at a time.

\item \textbf{Memory safety.} Rust's ownership model eliminates use-after-free bugs, data races, and buffer overflows at compile time---categories of bugs that have affected C++ numerical codes.

\item \textbf{Deployment simplicity.} ripopt compiles with \texttt{cargo build} with zero C or Fortran dependencies, eliminating the MUMPS/BLAS/LAPACK dependency chain that makes Ipopt installation non-trivial.
\end{enumerate}

\subsubsection{Architecture and Interface}
\label{org-architecture}
ripopt exposes a clean \texttt{NlpProblem} trait that users implement:

\begin{minted}[frame=lines,fontsize=\scriptsize,linenos]{rust}
pub trait NlpProblem {
    fn num_variables(&self) -> usize;
    fn num_constraints(&self) -> usize;
    fn bounds(&self, x_l: &mut [f64], x_u: &mut [f64]);
    fn constraint_bounds(&self, g_l: &mut [f64], g_u: &mut [f64]);
    fn initial_point(&self, x0: &mut [f64]);
    fn objective(&self, x: &[f64]) -> f64;
    fn gradient(&self, x: &[f64], grad: &mut [f64]);
    fn constraints(&self, x: &[f64], g: &mut [f64]);
    fn jacobian_structure(&self) -> (Vec<usize>, Vec<usize>);
    fn jacobian_values(&self, x: &[f64], vals: &mut [f64]);
    fn hessian_structure(&self) -> (Vec<usize>, Vec<usize>);
    fn hessian_values(&self, x: &[f64], obj_factor: f64,
                      lambda: &[f64], vals: &mut [f64]);
}
\end{minted}

The core IPM implements filter line search with Ipopt-style acceptance criteria, inertia correction via the \texttt{LinearSolver} trait (with dense LDL$^T$ and sparse LDL$^T$ backends implemented in Rust), second-order corrections, feasibility restoration, warm-starting, and configurable convergence criteria.

\subsubsection{Extensions Beyond Ipopt}
\label{org-ripopt-extensions}

ripopt extends Ipopt's algorithm in several significant ways:

\paragraph{Hybrid solver strategies.}
Rather than relying solely on the IPM, ripopt implements a multi-strategy fallback system that adapts to problem characteristics:
\begin{itemize}
\item \textbf{L-BFGS for unconstrained problems}: avoids pathological IPM behavior on unconstrained problems by using limited-memory BFGS with Wolfe line search, requiring only gradient evaluations.
\item \textbf{Augmented Lagrangian fallback}: when the IPM stalls on constrained problems, ripopt switches to a Powell--Hestenes--Rockafellar augmented Lagrangian method that solves a sequence of unconstrained subproblems via L-BFGS, with automatic penalty increase and multiplier updates.
\item \textbf{Slack variable reformulation}: converts inequality constraints to equalities with explicit slack variables, providing a cleaner KKT structure when the standard barrier approach struggles.
\end{itemize}

\paragraph{Automatic problem reformulation.}
ripopt detects overdetermined nonlinear equation systems ($m > n$, all equality constraints, infeasible starting point) and automatically reformulates them as least-squares problems $\min \frac{1}{2}\|g(x) - \text{target}\|^2$ with Gauss--Newton Hessian approximation ($H = J^T J$). This correctly classifies 99 overdetermined CUTEst problems as locally infeasible, avoiding the false convergence that a standard IPM would report.

\paragraph{Enhanced restoration phase.}
The feasibility restoration phase uses a three-stage fallback: (1) Gauss--Newton with adaptive Levenberg--Marquardt regularization, (2) penalty-regularized steps when Gauss--Newton fails, and (3) gradient descent as a last resort ($\Delta x = -J^T \cdot \text{violation}$). A watchdog strategy temporarily relaxes filter acceptance to escape narrow feasible corridors.

\paragraph{Native Rust linear algebra.}
Rather than depending on external Fortran libraries, ripopt implements dense LDL$^T$ factorization with Bunch--Kaufman pivoting and sparse LDL$^T$ factorization entirely in Rust, with native inertia detection from the diagonal of D. For problems with many more constraints than variables ($m \gg n$), a condensed KKT system using the Schur complement reduces factorization cost from $O((n+m)^3)$ to $O(n^2 m + n^3)$.

\paragraph{NLP scaling.}
Gradient-based objective scaling ensures the maximum gradient norm is $\le 100$ at the initial point, with per-constraint scaling matching Ipopt's \texttt{nlp\_scaling\_method = gradient-based}. This improves numerical stability without user intervention.

\subsubsection{Benchmark Results}
\label{org-ripopt-benchmarks}
On the 120-problem Hock--Schittkowski test suite, ripopt solves 119/120 (99.2\%), compared to 118/120 for Ipopt via cyipopt. On the 727-problem CUTEst test set, ripopt solves 583 (80.2\%) compared to Ipopt's 558 (76.8\%), with ripopt solving 25 more problems. Of the 544 problems both solvers solve, 452 agree to $10^{-4}$ relative tolerance. ripopt is substantially faster on small-to-medium problems: median speedup of 29.7$\times$ across the 548 commonly-solved problems, with 93\% of problems solved faster and 72\% solved 10$\times$+ faster.

\subsubsection{PyO3 Integration}
\label{org-pyo3-integration}
The \texttt{discopt-python} crate wraps ripopt with PyO3 bindings \cite{pyo3}. A \texttt{PyNlpProblem} struct implements the \texttt{NlpProblem} trait by calling back into Python evaluator methods (\texttt{evaluate\_objective}, \texttt{evaluate\_gradient}, \texttt{evaluate\_constraints}, \texttt{evaluate\_jacobian}, \texttt{evaluate\_lagrangian\_hessian}). Bounds and sparsity structure are cached in Rust-owned memory. The main solve loop releases the Python GIL via \texttt{py.allow\_threads()}, and callbacks re-acquire it only for the duration of each evaluation:


\begin{minted}[frame=lines,fontsize=\scriptsize,linenos]{rust}
let result = py.allow_threads(|| ripopt::solve(&problem, &opts));
\end{minted}

This means the heavy linear algebra in the IPM iteration runs without GIL contention, while the JAX-compiled gradient and Hessian evaluations use Python callbacks through the GIL.

\subsubsection{Licensing}
\label{org-licensing}
Because ripopt is a source-level translation of Ipopt (EPL-2.0), it is distributed under the Eclipse Public License 2.0. EPL-2.0 is not viral: code that \textit{depends on} ripopt does not need to adopt EPL-2.0, however, we have implemented a pure Python version in jax of the Ipopt algorithm in discopt, and have adopted the EPL-2.0 license for the entire discopt repository to avoid confusion and ensure that all components are covered by the same license.

\subsection{Ipopt via cyipopt}
\label{org-ipopt-via-cyipopt}
The third backend uses the industry-standard Ipopt solver \cite{Wachter2006} via cyipopt Python bindings. This serves as the correctness reference: any discrepancy between the pure-JAX IPM or ripopt and Ipopt on a well-conditioned problem indicates a bug in the former. Ipopt's advantage is its mature sparse linear algebra (MUMPS, MA27/MA57) which scales better on large, sparse problems ($n > 1000$). Its disadvantage is the Fortran/C++ dependency chain and inability to batch or differentiate through the solve.


\section{Convex Relaxation Framework}
\label{org-convex-relaxation-framework}
The quality of convex relaxations at B\&B nodes directly determines solver efficiency. discopt provides a compositional relaxation compiler that walks the expression DAG and produces JIT-compatible relaxation functions.

\subsection{McCormick Relaxation Compiler}
\label{org-mccormick-relaxation-compiler}
The relaxation compiler implements 25 univariate and bivariate McCormick relaxation functions \cite{McCormick1976,Tsoukalas2014} covering all operations supported by the modeling API: bilinear products, addition, subtraction, negation, division, power, \texttt{exp}, square, \texttt{abs}, \texttt{sqrt}, \texttt{log}, \texttt{log2}, \texttt{log10}, \texttt{sin}, \texttt{cos}, \texttt{tan}, \texttt{atan}, \texttt{asin}, \texttt{acos}, \texttt{sinh}, \texttt{cosh}, \texttt{tanh}, \texttt{sign}, \texttt{min}, and \texttt{max}.


Each function maps interval bounds $[x_l, x_u]$ and convex/concave enclosures $(cv, cc)$ of the argument to tighter enclosures of the result. The compiler produces pure \texttt{jax.numpy} functions compatible with \texttt{jax.jit} and \texttt{jax.vmap}, enabling batch relaxation evaluation over multiple B\&B nodes.

\subsection{$\alpha$ BB Underestimators}
\label{org-alpha-bb-underestimators}
For general twice-differentiable functions where McCormick relaxations are loose, the $\alpha$ BB method \cite{Androulakis1995,Adjiman1998} adds a separable quadratic perturbation:


\begin{equation}
L_{\alpha}(x) = f(x) - \sum_{i=1}^n \alpha_i (x_i - x_i^l)(x_i^u - x_i)
\end{equation}


The parameters $\alpha_i$ are chosen such that $L_\alpha$ is convex: $\alpha_i \ge \max(0, -\lambda_{\min}(\nabla^2 f) / 2)$ where $\lambda_{\min}$ is the minimum eigenvalue of the Hessian. discopt provides two estimation methods:


\begin{itemize}
\item \textbf{Eigenvalue method}: exact computation via \texttt{jax.hessian} and \texttt{jnp.linalg.eigvalsh} at sampled points (accurate but requires $O(n^3)$ per sample)

\item \textbf{Gershgorin method}: conservative estimate from Hessian diagonal dominance (faster, slightly looser bounds, zero soundness violations in testing)

\end{itemize}


The JAX-native $\alpha$ BB estimation uses \texttt{jax.vmap} over sample points, avoiding the finite-difference fallback required for \texttt{.nl} file models by leveraging exact Hessians from the expression DAG.

\subsection{Piecewise McCormick Relaxations}
\label{org-piecewise-mccormick-relaxations}
Piecewise McCormick \cite{Castro2015} partitions the variable domain into $k$ subintervals and constructs tighter McCormick envelopes on each partition. discopt supports both uniform partitioning and adaptive partitioning that concentrates breakpoints where $f''(x)$ is largest.


On representative test functions, piecewise McCormick with $k=8$ partitions achieves significant gap reduction versus standard McCormick:


\begin{table}[htbp]
\centering
\begin{tabular}{llll}
\toprule
Function, interval & Standard gap & Piecewise gap & Reduction \\
\midrule
$\exp(x)$, $[-2,3]$ & 6.180 & 0.129 & 97.9\% \\
$\log(x)$, $[0.5,8]$ & 0.566 & 0.017 & 97.0\% \\
$x^2$, $[-3,4]$ & 8.098 & 0.127 & 98.4\% \\
bilinear $xy$ & varies & varies & >30\% \\
\bottomrule
\end{tabular}
\caption{Piecewise McCormick gap reduction with $k=8$ uniform partitions. Gap = mean(cc $-$ cv) over 10,000 random sample points. Results reproduced from \texttt{piecewise\_gap\_reduction.ipynb}.}
\label{tab-piecewise}
\end{table}

\subsection{Convexity Detection}
\label{org-convexity-detection}
An expression DAG convexity detector classifies each subexpression as \textsc{affine}, \textsc{convex}, \textsc{concave}, or \textsc{unknown} using standard composition rules from convex analysis \cite{Boyd2004}. The key rules are:

\begin{itemize}
\item Constants, variables, and parameters are \textsc{affine}.
\item Sum of two \textsc{convex} (resp.~\textsc{concave}) expressions preserves curvature; \textsc{convex} + \textsc{concave} yields \textsc{unknown}.
\item A positive constant times a \textsc{convex} expression is \textsc{convex}; a negative constant flips curvature.
\item $\exp(\text{convex})$ is \textsc{convex} (composition with nondecreasing convex outer function).
\item $\log(\text{concave})$ is \textsc{concave} (composition with nondecreasing concave outer function).
\item $x^{2k}$ (even integer power of affine) is \textsc{convex}; $x^p$ for $p > 1$ on nonneg domain is \textsc{convex}; $x^p$ for $0 < p < 1$ on nonneg domain is \textsc{concave}.
\item Bilinear terms $x \cdot y$ are \textsc{unknown}.
\end{itemize}

Results are cached on expression IDs for efficiency. The detector runs in time linear in the DAG size and provides three downstream optimizations:

\begin{enumerate}
\item \textbf{Adaptive solver dispatch (E2):} When the entire model is classified as convex (convex objective for minimization, all constraints convex), the solver skips $\alpha$BB convexification and McCormick relaxation compilation entirely. NLP solutions are directly valid lower bounds, eliminating relaxation overhead.

\item \textbf{Per-constraint OA cuts (E3):} In a nonconvex problem, individual constraints may still be convex. The convexity mask enables outer approximation cuts \cite{Duran1986} on a per-constraint basis: tangent hyperplanes are globally valid underestimators for convex constraints, even when the overall problem is nonconvex. This replaces the earlier approach that only generated OA for affine (linear) constraints.

\item \textbf{Constraint classification:} A constraint $g(x) \le 0$ is convex when $g$ is \textsc{convex}; $g(x) \ge 0$ is convex when $g$ is \textsc{concave}; $g(x) = 0$ is convex only when $g$ is \textsc{affine}. The model-level classification checks the objective and all constraints simultaneously.
\end{enumerate}

\subsection{Specialized Envelopes}
\label{org-specialized-envelopes}
Standard McCormick relaxations compose relaxations through the expression tree, which can introduce composition error: outer functions see the relaxation of inner functions rather than the original function. Specialized envelopes compute exact convex/concave envelopes directly for common operations, bypassing this composition gap.

discopt provides the following envelope functions:
\begin{itemize}
\item \texttt{relax\_power\_int($x, l, u, p$)}: exact envelope for $x^p$ with integer $p$. For even $p$, the function is convex everywhere, so the convex envelope is $x^p$ itself and the concave envelope is the secant line. For odd $p \ge 3$, convexity/concavity depends on sign: convex on $[0, \infty)$, concave on $(-\infty, 0]$, and for mixed-sign domains a tangent line construction provides tight bounds.
\item \texttt{relax\_exp\_tight($x, l, u$)}: convex envelope $= \exp(x)$, concave envelope $= $ secant line from $(l, e^l)$ to $(u, e^u)$.
\item \texttt{relax\_log\_tight($x, l, u$)}: concave envelope $= \log(x)$, convex envelope $= $ secant line.
\item \texttt{relax\_exp\_bilinear}, \texttt{relax\_log\_sum}, \texttt{relax\_trilinear}, \texttt{relax\_signomial}: envelopes for composite operations.
\end{itemize}

\paragraph{Tight trigonometric envelopes.}
For $\sin(x)$ on a bounded interval $[l, u]$, the relaxation quality depends on the curvature regime \cite{Locatelli2014}. \texttt{relax\_sin\_tight} identifies three cases:
\begin{itemize}
\item \emph{Concave regime} ($\sin'' < 0$, interval within $[2k\pi, (2k+1)\pi]$): cv $=$ secant, cc $= \sin(x)$.
\item \emph{Convex regime} ($\sin'' > 0$, interval within $[(2k-1)\pi, 2k\pi]$): cv $= \sin(x)$, cc $=$ secant.
\item \emph{Mixed or wide} ($\ge 2\pi$): falls back to $[-1, 1]$.
\end{itemize}
\texttt{relax\_cos\_tight} delegates via $\cos(x) = \sin(x + \pi/2)$. The compiler auto-dispatches these when the argument is a plain variable with known bounds.

\paragraph{Multivariate signomial envelopes.}
For signomial terms $\prod_{i=1}^n x_i^{a_i}$ with positive variables, \texttt{relax\_signomial\_multi} decomposes the product via logarithms \cite{Maranas1997}:
$\prod x_i^{a_i} = \exp\!\bigl(\sum a_i \log x_i\bigr)$.
Each $\log x_i$ is relaxed using tight log envelopes (concave), scaled by the exponent (sign-aware), summed, and composed with \texttt{exp} (convex). The compiler detects products of $\text{Variable}^{\text{Constant}}$ terms and routes to this envelope when all lower bounds are positive.

\paragraph{Additional univariate envelopes.}
The envelope library also provides relaxations for six additional operations, each exploiting known curvature properties:
\begin{itemize}
\item \texttt{relax\_asinh}: $\operatorname{asinh}''(x) = -x/(1+x^2)^{3/2}$; concave on $[0,\infty)$, convex on $(-\infty,0]$.
\item \texttt{relax\_acosh}: concave on $[1,\infty)$.
\item \texttt{relax\_atanh}: $\operatorname{atanh}''(x) = 2x/(1-x^2)^2$; convex on $[0,1)$, concave on $(-1,0]$.
\item \texttt{relax\_erf}: $\operatorname{erf}''(x) = -\frac{4x}{\sqrt{\pi}} e^{-x^2}$; convex on $(-\infty,0]$, concave on $[0,\infty)$.
\item \texttt{relax\_log1p}: $\log(1+x)$ is concave on $(-1,\infty)$; numerically stable near $x=0$.
\item \texttt{relax\_reciprocal}: $1/x$ is convex on $(0,\infty)$, concave on $(-\infty,0)$.
\end{itemize}
For each, the convex regime uses function-value underestimator and secant overestimator, and vice versa for the concave regime. Mixed-sign domains use conservative min/max combinations.

The relaxation compiler automatically selects specialized envelopes when the expression matches a known pattern---for example, when $x^p$ is encountered and $x$ is a variable (or indexed variable) with known bounds, \texttt{relax\_power\_int} is used with the actual variable bounds from the B\&B node rather than propagated interval bounds. This provides tighter relaxations because the envelope is exact for the monomial, whereas McCormick composition of $(x^2)^2$ for $x^4$ introduces two layers of relaxation error.

\subsection{Learned Relaxations via ICNN}
\label{org-learned-relaxations-via-icnn}
As an experimental feature, discopt supports learned convex relaxations using input convex neural networks (ICNNs) \cite{Amos2017icnn}. An ICNN architecture with softplus non-negative weight constraints is trained to approximate tighter-than-McCormick underestimators for specific operations, with a soundness loss term that penalizes violations of the convexity and underestimation properties.


A \texttt{LearnedRelaxationRegistry} provides per-operation wrappers with automatic fallback to standard McCormick for unsupported operations. In testing, ICNN relaxations achieve 5--10\% gap reduction beyond standard McCormick with lightweight training (200 epochs, 5k boxes), with the potential for 30\%+ with more extensive training (500 epochs, 50k boxes).


\section{Differentiable Optimization}
\label{org-differentiable-optimization}
A distinguishing feature of discopt is the ability to differentiate through the optimization solve with respect to problem parameters. This enables decision-focused learning \cite{Elmachtoub2022} where the model parameters (e.g., cost coefficients) are learned end-to-end from decision outcomes rather than prediction accuracy.

\subsection{Level 1: Envelope Theorem}
\label{org-level-1-envelope-theorem}
For a parametric NLP $\min_x f(x; p)$ subject to $g(x; p) \le 0$, the envelope theorem gives:


\begin{equation}
\frac{d f^*}{d p} = \frac{\partial \mathcal{L}}{\partial p}\bigg|_{x^*, \lambda^*} = \frac{\partial f}{\partial p}\bigg|_{x^*} + {\lambda^*}^T \frac{\partial g}{\partial p}\bigg|_{x^*}
\end{equation}


This requires no additional linear system solve---the optimal dual variables $\lambda^*$ from the NLP solver directly provide the sensitivity. discopt implements this as a \texttt{DiffSolveResult.gradient(param)} method that returns $df^*/dp$ for any model parameter.

\subsection{Level 3: Implicit Differentiation via KKT}
\label{org-level-3-implicit-differentiation-via-kkt}
For the full primal sensitivity matrix $dx^*/dp$, discopt differentiates the KKT conditions:


\begin{equation}
\begin{bmatrix}
\nabla^2_{xx}\mathcal{L} & J_a^T \\
J_a & 0
\end{bmatrix}
\begin{bmatrix}
dx^*/dp \\
d\lambda^*/dp
\end{bmatrix}
=
-\begin{bmatrix}
\nabla^2_{xp}\mathcal{L} \\
\partial g_a / \partial p
\end{bmatrix}
\end{equation}


where $J_a$ is the Jacobian of active constraints (including active variable bounds) and $\nabla^2_{xp}\mathcal{L}$ is the mixed Hessian of the Lagrangian. This is computed via \texttt{jax.hessian} and \texttt{jax.jacobian} applied to parametric DAG-compiled functions.


The \texttt{DiffSolveResultL3} class provides:


\begin{itemize}
\item \texttt{sensitivity\_matrix()}: full $dx^*/dp$ matrix

\item \texttt{implicit\_gradient(param)}: $df^*/dp$ via the chain rule $\nabla_x f \cdot dx^*/dp + \partial f/\partial p$

\item \texttt{approximate\_resolve(new\_params)}: first-order approximation $x^* + (dx^*/dp) \Delta p$

\item \texttt{dual\_sensitivity(param)}: $d\lambda^*/dp$ for constraint multiplier sensitivity

\item \texttt{reduced\_hessian()}: Hessian projected onto the null space of active constraints (for uncertainty quantification)

\item \texttt{sensitivity(dp)}: fast re-query with cached KKT factorization (analogous to sIPOPT \cite{Pirnay2012})

\end{itemize}


If the KKT system is ill-conditioned, discopt falls back to finite-perturbation smoothing, and then to L1 envelope theorem gradients.

\subsection{Application: Decision-Focused Learning}
\label{org-application-decision-focused-learning}
The JAX-native differentiable solve is exposed via \texttt{jax.custom\_jvp}, enabling it to be embedded in a larger JAX computation graph. We demonstrate decision-focused learning on a synthetic resource allocation task with a linear prediction model $\hat{c} = \theta \phi$, where $\theta \in \mathbb{R}^{3 \times 5}$ is a weight matrix and $\phi \in \mathbb{R}^5$ is a feature vector. The true cost function includes nonlinear interaction terms that the linear model cannot capture, creating systematic prediction errors. The pipeline proceeds as follows:

\begin{enumerate}
\item The linear model $\hat{c} = \theta \phi$ predicts cost parameters from features

\item The predicted costs are fed to a risk-averse allocation QP: $\min_x \hat{c}^T x + \varepsilon \|x\|^2$ subject to $\sum x_i = 1$, $x_i \in [0, 1]$, solved via \texttt{discopt.solve()} to obtain $x^*(\hat{c})$

\item The true task loss $\ell = c_{\mathrm{true}}^T x^*(\hat{c})$ is computed using the actual (observed) costs

\item The sensitivity matrix $dx^*/d\hat{c}$ is obtained via L3 implicit differentiation of the KKT conditions (\texttt{differentiable\_solve\_l3})

\item The chain rule gives $\partial \ell / \partial \theta = (dx^*/d\hat{c})^T c_{\mathrm{true}} \otimes \phi$, and $\theta$ is updated via gradient descent to minimize decision regret
\end{enumerate}

The quadratic regularization $\varepsilon = 1.0$ is essential: it keeps the optimal allocation in the interior of $[0, 1]$, ensuring the KKT system is non-degenerate and the sensitivity matrix $dx^*/d\hat{c}$ is non-trivial. Without sufficient regularization, the solution sits at a vertex where $dx^*/d\hat{c} = 0$ and no gradient flows through the solver.

On a resource allocation task with 3 activities, 5 features, and 30 training samples (single-seed demonstration with \texttt{np.random.seed(42)}), decision-focused learning achieves \textbf{36\% regret reduction} compared to two-stage predict-then-optimize (where $\theta$ is trained with MSE via ordinary least squares), despite having higher prediction MSE. This confirms the key insight of decision-focused learning \cite{Elmachtoub2022}: optimizing for decision quality beats optimizing for prediction accuracy when prediction errors are systematic. On a mean-variance portfolio optimization task with 3 assets and nonlinear interactions, DFL similarly achieves lower regret and higher utility than the two-stage approach. Both experiments use a linear predictor; extending to neural network predictors (e.g., via Equinox or Flax) is a natural next step that would allow capturing the nonlinear cost dependencies directly.


\section{Computational Experiments}
\label{org-computational-experiments}
\subsection{NLP Backend Comparison}
\label{org-nlp-backend-comparison}
We compare the three NLP backends on 9 test problems spanning unconstrained quadratics to nonlinear constraints with trigonometric, exponential, logarithmic, and bilinear terms. All problems have $n = 2$ variables; the number of constraints $m$ ranges from 0 to 2. Table~\ref{tab-nlp-backends} reports objectives and wall-clock timing. All timings were collected on an Apple M4 Pro (CPU backend, JAX 0.8.2) and represent single-solve wall-clock time including model compilation. Results are reproducible from the \texttt{ipm\_vs\_ipopt.ipynb} notebook.


\begin{table}[htbp]
\centering
\begin{tabular}{llrrrr}
\toprule
Problem & $m$ & $f^*$ & IPM (s) & ripopt (s) & Ipopt (s) \\
\midrule
Unconstrained quadratic & 0 & 0.000 & 1.029 & 0.011 & 0.009 \\
Rosenbrock & 0 & 0.000 & 0.462 & 0.075 & 0.106 \\
Constrained quadratic & 1 & 0.500 & 0.164 & 0.001 & 0.000 \\
Nonlinear equality ($xy = 1$) & 1 & 2.000 & 0.349 & 0.114 & 0.091 \\
Exponential NLP & 1 & 1.839 & 0.261 & 0.094 & 0.091 \\
Logarithmic NLP & 1 & 1.386 & 0.252 & 0.087 & 0.091 \\
Multiple constraints & 1 & 0.500 & 0.103 & 0.014 & 0.012 \\
Bilinear constraints & 2 & 2.000 & 0.318 & 0.105 & 0.132 \\
Trigonometric NLP & 1 & $-0.995$ & 0.264 & 0.094 & 0.093 \\
\midrule
Total (9 problems) &  &  & 3.20 & 0.59 & 0.63 \\
\bottomrule
\end{tabular}
\caption{NLP backend comparison on 9 test problems (all $n = 2$). All backends achieve identical objectives (to solver tolerance $10^{-6}$). Timings on Apple M4 Pro, CPU. IPM times include one-time JIT compilation.}
\label{tab-nlp-backends}
\end{table}


Key observations:

\begin{itemize}
\item All three backends produce \textbf{identical optimal objectives} to solver tolerance ($10^{-6}$) on all 9 problems, confirming correctness across the Rust, JAX, and C++ implementations.

\item ripopt is competitive with Ipopt (total 0.59s vs 0.63s), and faster on Rosenbrock (0.075s vs 0.106s) and bilinear constraints (0.105s vs 0.132s).

\item The pure-JAX IPM times are dominated by one-time JIT compilation overhead (XLA tracing and optimization). Subsequent solves of similarly-structured problems reuse cached JIT traces and are substantially faster. The IPM total (3.20s) reflects 9 cold-start compilations; in a B\&B loop where the same compiled function is called thousands of times, per-solve cost drops to the millisecond range.
\end{itemize}

\subsection{MINLP Correctness}
\label{org-minlp-correctness}
We validate on 24 MINLP problems (12 NLP + 12 MINLP with binary/integer variables) from the discopt test suite, with known optimal values:


\begin{itemize}
\item \textbf{24/24 solved correctly} with zero incorrect results

\item Tolerances: absolute $10^{-4}$, relative $10^{-3}$ for objectives; integrality tolerance $10^{-5}$

\item Problems include Rosenbrock, exponential, logarithmic, trigonometric, bilinear, and cardinality-constrained formulations

\end{itemize}


B\&B convergence with the ripopt backend was verified on two MINLP instances: a simple 3-variable problem (optimal $f^* = 2.0$, 1 node, 0.039s) and a larger 6-variable formulation (optimal $f^* = 3.0$, 13 nodes, 2.38s).

\subsubsection{MINLPLib Validation}
\label{org-minlplib-validation}
We further validate on instances from MINLPLib \cite{Bussieck2003}, imported via the \texttt{.nl} file parser. On 73 instances with $\le 50$ variables, discopt solves 49 to optimal/feasible status, with 18 of 27 checked instances matching known optima. Table \ref{tab-minlplib} reports results on a representative 9-instance smoke test suite.


\begin{table}[htbp]
\centering
\begin{tabular}{llllllll}
\toprule
Instance & $n$ & $m$ & Status & $f^*$ & Best Known & Time (s) & Nodes \\
\midrule
chance & 4 & 3 & iter.~limit & --- & 29.894 & 0.87 & 0 \\
dispatch & 4 & 2 & iter.~limit & --- & 3155.288 & 0.72 & 0 \\
ex1221 & 5 & 5 & optimal & 7.667 & 7.667 & 2.84 & 1 \\
ex1226 & 5 & 5 & optimal & $-$17.000 & $-$17.000 & 7.29 & 3 \\
gear & 4 & 0 & feasible & 0.000 & 0.000 & 73.87 & 93 \\
nvs01 & 3 & 3 & feasible & 0.034 & 12.470 & 117.52 & 53 \\
nvs03 & 2 & 2 & optimal & 16.000 & 16.000 & 4.02 & 7 \\
nvs04 & 2 & 0 & optimal & 0.720 & 0.720 & 43.76 & 59 \\
nvs06 & 2 & 0 & optimal & 1.770 & 1.770 & 39.95 & 9 \\
\bottomrule
\end{tabular}
\caption{MINLPLib smoke test results (9 instances, 60s time limit). Best known objectives from MINLPLib. Results reproduced from \texttt{minlplib\_smoke\_test.ipynb}.}
\label{tab-minlplib}
\end{table}


Of the 9 instances, 4 reach optimal status (matching known optima to $10^{-3}$), 3 reach feasible solutions, and 2 hit the iteration limit without converging. The \texttt{gear} instance finds a near-zero objective (matching the known optimum $0.0$) but cannot prove optimality without convex relaxations for the \texttt{.nl} import path. Instances \texttt{chance} and \texttt{dispatch} fail to converge due to non-convex NLP structure. The \texttt{.nl} import path currently lacks the expression-DAG-level relaxation support available to native discopt models, which limits both bound tightening and optimality proof on these instances.

\subsection{Relaxation Gap Reduction}
\label{org-relaxation-gap-reduction}
Table \ref{tab-piecewise} reports piecewise McCormick gap reduction. Additionally:


\begin{itemize}
\item $\alpha$ BB underestimators achieve zero soundness violations across 1000+ test points with both eigenvalue and Gershgorin methods

\item ICNN learned relaxations provide 5--10\% additional gap reduction beyond standard McCormick with lightweight training

\item The relaxation compiler supports mode switching between \texttt{standard}, \texttt{piecewise}, and \texttt{learned} at the solver level

\end{itemize}

\subsection{Algebraic Coefficient Extraction}
\label{org-performance}
For LP and QP subproblems, discopt walks the expression DAG algebraically to extract coefficient matrices ($c$, $A$, $b$ for LP; additionally $Q$ for QP), bypassing JAX tracing and automatic differentiation entirely. On random instances with $n = 100$ variables and $m = 50$ constraints:

\begin{itemize}
\item LP extraction: 0.024s algebraic vs.\ 4.93s autodiff (\textbf{209$\times$} speedup)
\item QP extraction: 0.003s algebraic vs.\ 8.32s autodiff (\textbf{2539$\times$} speedup)
\item Algebraic end-to-end solve: 0.013s (LP warm), 0.017s (QP warm)
\end{itemize}

The algebraic path also produces correct solutions (optimal status) on problems where the generic NLP path fails to converge within iteration limits. Results reproduced from \texttt{benchmark\_lp\_qp\_speedup.ipynb}.


\section{Discussion}
\label{org-discussion}
discopt demonstrates that a modern, multi-language MINLP solver can provide capabilities absent from existing monolithic solvers: differentiability, batch evaluation, and memory safety. Several limitations and directions for future work deserve discussion.

\textbf{Convexity detection and adaptive dispatch.} The expression DAG convexity detector implements the Disciplined Convex Programming (DCP) composition rules \cite{Boyd2004}, covering the operations supported by the modeling API. When a model is fully convex, the solver bypasses $\alpha$BB convexification and McCormick relaxation compilation, relying instead on the NLP solution as a valid lower bound. For partially convex models, per-constraint classification enables outer approximation cuts on convex constraints while leaving nonconvex constraints to be handled by spatial branching. The current detector is conservative: expressions not covered by the composition rules are classified as \textsc{unknown} even if they happen to be convex, and the detector does not exploit second-order information (Hessian positive-semidefiniteness checks) for numerical verification.

\textbf{Branching and bound tightening.} Reliability branching \cite{Achterberg2005} with LP-based strong branching provides a practical variable selection scheme that approaches full strong branching quality at amortized low cost. The pseudocost tracking in Rust provides per-variable observation counts that trigger strong branching only for unreliable variables. OBBT with incumbent cutoff \cite{Gleixner2017} progressively tightens variable bounds throughout the tree, with each bound improvement propagating to all open nodes. The combination of reliability branching and incumbent-based OBBT addresses two of the key algorithmic gaps relative to BARON \cite{Tawarmalani2005}.

\textbf{Specialized envelopes versus compositional relaxations.} The envelope functions for power, exponential, logarithmic, trigonometric, and signomial operations provide exact convex/concave envelopes, avoiding the composition error inherent in McCormick's recursive approach. The relaxation compiler automatically detects when a specialized envelope is applicable and routes accordingly. Recent additions include regime-aware $\sin$/$\cos$ envelopes \cite{Locatelli2014}, multivariate signomial relaxations via logarithmic decomposition \cite{Maranas1997}, and envelopes for $\operatorname{asinh}$, $\operatorname{acosh}$, $\operatorname{atanh}$, $\operatorname{erf}$, $\log(1+x)$, and $1/x$ that exploit known curvature properties.

\textbf{Generalized disjunctive programming.} The GDP reformulation layer \cite{Raman1994,Grossmann2002} enables higher-level modeling of discrete--continuous problems with indicator constraints, disjunctions, and logical propositions. The automatic big-M computation via interval arithmetic typically produces tighter M values than manual specification, and the convex hull reformulation \cite{Lee2001} provides provably tighter LP relaxations at the cost of additional disaggregated variables. The perspective function formulation for nonlinear constraints in the hull reformulation uses $\epsilon$-clamping for numerical stability; tighter formulations based on the extended perspective function \cite{Trespalacios2015} are a direction for future work.

\textbf{Limitations.} The solver's performance on large-scale problems ($n > 1000$) is limited by the dense linear algebra in the pure-JAX IPM; the sparse IPM variant using lineax iterative solvers partially addresses this but lacks the numerical robustness of direct factorization methods. The GNN branching model currently runs with random weights; training on problem-specific data would improve variable selection. Head-to-head benchmarking against BARON on MINLPLib instances remains future work pending license acquisition.


\section{Conclusion}
\label{org-conclusion}
We have presented discopt, a hybrid Rust/JAX/Python MINLP solver that bridges the gap between classical mathematical programming and modern differentiable programming. The solver provides three interchangeable NLP backends---a pure-JAX IPM, ripopt (a memory-safe Rust translation of Ipopt), and Ipopt via cyipopt---all producing identical results on benchmark problems. The differentiable solving capability enables decision-focused learning through nonconvex optimization, achieving 36\% regret reduction on resource allocation tasks. The compositional relaxation compiler achieves 97--98\% gap reduction via piecewise McCormick, and specialized envelopes for power, exponential, logarithmic, trigonometric, signomial, and inverse-hyperbolic operations provide exact convex/concave envelopes that avoid compositional relaxation error. A generalized disjunctive programming layer supports indicator constraints, disjunctions with big-M and convex hull reformulations, SOS constraints, and logical propositions, enabling natural modeling of discrete--continuous problems. An expression DAG convexity detector enables adaptive solver dispatch (skipping relaxation overhead for convex problems) and per-constraint outer approximation cuts. Reliability branching with pseudocost tracking and LP-based strong branching provides near-optimal variable selection, while OBBT with incumbent cutoff progressively tightens bounds throughout the search tree. The algebraic coefficient extraction path provides 209--2539$\times$ speedups over autodiff-based extraction for LP and QP subproblems. ripopt eliminates the Fortran/C++ dependency chain while preserving Ipopt's numerical robustness and extending it with hybrid fallback strategies (L-BFGS, augmented Lagrangian), automatic least-squares reformulation, and native Rust linear algebra, achieving 99.2\% solve rate on the Hock--Schittkowski suite and 80.2\% on the 727-problem CUTEst suite (vs.\ Ipopt's 76.8\%), with a median 29.7$\times$ speedup.


The system is open-source: discopt and ripopt are available under the EPL-2.0 License.


\bibliographystyle{unsrtnat}
\bibliography{references}


\end{document}